How to Avoid OOTA Without Really Trying

### 2024-09-02 paper/ootavacuous.tex
ootavacuous: Earlier self-review and taglines
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

This paper shows that compiler-based C++ implementations subject to
reasonable constraints on how they treat accesses to atomic objects
cannot exhibit out-of-thin-air (OOTA) cycles.
It follows that these implementations need to take almost no special
actions to avoid OOTA.
In fact, for many compilers the constraints boil down to a simple
``Don't invent or duplicate atomic accesses'', which the compiler
probably wouldn't do anyway.
Absolutely no changes are required to either the standard or to user code.

@@

This paper will follow P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
in using the term \emph{full C++} to denote the standard including the
prohibition of OOTA mentioned above.
Unlike that article, we will use the term \emph{loose C++}
(rather than \emph{strict C++}, which seems too similar to
\emph{full C++} for comfort)
to denote a hypothetical standard that excludes this prohibition but
is otherwise identical to full C++.
Unqualified \emph{C++} means full C++.

@@

Other workers recommend various procedures to identify and avoid OOTA
cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
but none of these have been looked upon favorably by C++ implementers.
Some of these workers appear to have abandoned this effort, but as of
2024, Mark Batty is persisting with modular relaxed dependencies.
@@@ cite web site.

@@

@@

@@

@@

Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
page~\pageref{lst:Simple OOTA},
in which \co{thread1()} loads the value of \co{x} and stores it in
\co{y} while \co{thread2()} does the reverse.
A globally optimizing loose C++ compiler given that program might
transform it to the following before translating it into machine code,
if the compiler is sufficiently perverse:
\begin{quote}
\begin{verbatim}
 1 atomic<int> x(0);
 2 atomic<int> y(0);
 3
 4 void thread1()
 5 {
 6   int r1 = 42;
 7   y.store(r1, memory_order_relaxed);
 8 }
 9
10 void thread2()
11 {
12   int r2 = 42;
13   x.store(r2, memory_order_relaxed);
14 }
\end{verbatim}
\end{quote}
The loads previously on lines~6 and~12 have been replaced by constants.
Such a transformation complies with the loose C++ standard,
even though the resulting executable file would produce an unintuitive
OOTA outcome every time it runs!

@@

@@

@@

With the contents of the last few sections under our belts, we can
formulate the Fundamental Property that we would like all semantic
dependencies to satisfy.
Note that this formulation makes sense only for implementations in
which all atomic objects are treated as volatile or quasi volatile.
\begin{quote}
Let $W$ be a store which semantically depends on loads
$\{R_0$, \ldots,~$R_n\}$ in some abstract execution $A$,
and suppose that $W$ is not
omitted in some hardware execution $H$ realizing $A$.
Then for some $i$, load $R_i$ is not omitted in $H$ and the
instruction it maps to is ordered before the instruction $W$ maps to.
\end{quote}
It's quite straightforward to show that under any implementation in which
semantic dependencies satisfy the Fundamental Property, no OOTA cycle
has a nontrivial realization.

Indeed, suppose that abstract execution $A$ is realized by hardware
execution $H$ and $A$ has an OOTA cycle.
This means there are atomic stores $W_i$ in $A$,
semantically depending on atomic loads
$\{R_{i,j}\}$ where each of the loads reads from one of the $W_k$
stores in a different thread.
Let $W_i'$ and $R_{i,j}'$ be the hardware instructions these accesses
map to in $H$, if they aren't omitted.
Assuming that the stores are not all omitted,
one of the $W_i'$ instructions, let's say $W_0'$, must commit first.
By the Fundamental Property, one of the loads that $W_0$ depends on,
let's say $R_{0,0}$, is not omitted and $R_{0,0}'$ is ordered before $W_0'$.
But now we have a contradiction:
\begin{itemize}
\item	$W_0'$ commits after $R_{0,0}'$;
\item	$R_{0,0}'$ commits after the store instruction $W_k'$ it reads from;
\item	$W_k'$ commits no earlier than $W_0'$.
\end{itemize}

@@

>>>>

This paper shows that compiler-based C++ implementations subject to
reasonable constraints on how they treat accesses to atomic objects
cannot exhibit out-of-thin-air (OOTA) cycles.
It follows that these implementations need to take almost no special
actions to avoid OOTA.
In fact, for many compilers the constraints boil down to a simple ``Don't
invent or duplicate atomic accesses'', which the compiler should not
do anyway if it is to evaluate arithmetic expressions involving atomic
accesses in accordance with the abstract machine.
Absolutely no changes are required to either the standard or to user code.

@@

This paper follows P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
in using \emph{full C++} to denote the standard including the
prohibition of OOTA mentioned above.
Unlike that article, we use \emph{loose C++}
(rather than \emph{strict C++}, which seems too similar to
\emph{full C++} for comfort)
to denote a hypothetical standard that excludes this prohibition but
is otherwise identical to full C++.
Unqualified \emph{C++} means full C++.

@@

Other workers recommend various procedures to identify and avoid OOTA
cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
but none of these have been looked upon favorably by C++ implementers.
Some of these workers appear to have abandoned this effort, but as of
2024, Mark Batty is persisting with modular relaxed dependencies, which
may have the added ability of checking compiler optimizations.
@@@ cite web site.

@@

\emph{Tagline: ``Semantic dependencies are more complicated than you think.''}

% @@@
% 1: rfe, not rf.
% 2: Rules of arithmetic (*0).
% 3: Many-to-one sdep (*0 + ... )
% 4: Compiler knowledge, cross-thread special case.
% 5: Different stores.
% 6: "If" statements.
% 7: Matching up stores.
% Semantic dependencies depend on execution.
% @@@

@@

\emph{Tagline: ``Need to look at machine code.''}

@@

\emph{Tagline: Compilers are constrained in optimizing atomics, especially
volatile atomics.}

@@

% @@@ Needed?  If so, can this be in sdep section?

Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
page~\pageref{lst:Simple OOTA},
in which \co{thread1()} loads the value of \co{x} and stores it in
\co{y} while \co{thread2()} does the reverse.
A sufficiently perverse globally optimizing loose C++ compiler might
transform that program to the following before translating it into
machine code: \\
% Constructed by hand.
{
\small
\begin{tabular}{r|l|l}
% & \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
% \hline
1. &
\texttt{~~int r1 = 42} &
	\texttt{~~int r2 = 42} \\
2. &
\texttt{~~y =\textsubscript{rlx} r1} &
	\texttt{~~z =\textsubscript{rlx} r2} \\
\end{tabular}
} \\
The loads previously on lines~6 and~12 have been replaced by constants.
Such a transformation complies with the loose C++ standard,
even though the resulting executable file would produce an unintuitive
OOTA outcome every time it runs!

@@

\emph{ @@@ Merge into earlier discussion of inventing atomic loads. }

@@

\emph{Tagline: ``How to think about semantic dependencies.'' }

@@

With the contents of the last few sections under our belts, we can
formulate the Fundamental Property that we would like all semantic
dependencies to satisfy.
Note that this formulation makes sense only for implementations in
which all atomic objects are treated as volatile or quasi volatile.
\begin{quote}
Let $W$ be a store which semantically depends on loads
$\{R_0$, \ldots,~$R_n\}$ in some abstract execution $A$,
and suppose that $W$ is not
omitted in some hardware execution $H$ realizing $A$.
Then for some $i$, load $R_i$ is not omitted in $H$ and the
instruction it maps to is ordered before the instruction $W$ maps to.
\end{quote}

No implementation in which semantic dependencies satisfy the Fundamental
Property has a nontrivial OOTA cycle.
To see this, suppose that abstract execution $A$ is realized by hardware
execution $H$ and $A$ does have an OOTA cycle.
This means there are atomic stores $W_i$ in $A$,
semantically depending on atomic loads
$\{R_{i,j}\}$ where each of the loads reads from one of the $W_k$
stores in a different thread.
Let $W_i'$ and $R_{i,j}'$ be the hardware instructions these accesses
map to in $H$, if they aren't omitted.
Assuming that the stores are not all omitted,
one of the $W_i'$ instructions, let's say $W_0'$, must commit first.
By the Fundamental Property, one of the loads that $W_0$ depends on,
let's say $R_{0,0}$, is not omitted and $R_{0,0}'$ is ordered before $W_0'$.
But now we have a contradiction:
(1)~$W_0'$ commits after $R_{0,0}'$;
(2)~$R_{0,0}'$ commits after the store instruction $W_k'$ it reads from;
and finally
(3)~$W_k'$ commits no earlier than $W_0'$.

@@

\emph{ Tagline: ``How to think about OOTA.'' }
